---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/stan-dev/posteriordb-r/actions/workflows/check-release.yaml/badge.svg)](https://github.com/stan-dev/posteriordb-r/actions/workflows/check-release.yaml) [![Codecov test coverage](https://codecov.io/gh/stan-dev/posteriordb-r/branch/main/graph/badge.svg)](https://codecov.io/gh/stan-dev/posteriordb-r?branch=main)
<!-- badges: end -->

# `posteriordb`: an R package to work with `posteriordb`

This repository contains the R package to efficiently work with the [posteriordb](https://github.com/stan-dev/posteriordb) repository. The R package includes convenience functions to access data, model code and information for individual posteriors, models, data and draws.


## Installation

To install only the R package and then access the posteriors remotely, install the package from GitHub using the `remotes` package.

```{r, eval = FALSE}
remotes::install_github("stan-dev/posteriordb-r")
```

To load the package, just run.

```{r}
library(posteriordb)
```

## Connect to the posterior database

First, we create the posterior database connection to use. Here we want to use the database locally. We assume the `posteriordb` repo has been cloned and is accessible locally.

```{r, eval=FALSE}
my_pdb <- pdb_local()
```

Pass the database directory explicitly, or set `PDB_PATH`. The
`pdb_path` option takes precedence over that environment variable. The path
must contain the database's `data`, `models` and `posteriors` directories;
it may be the `posterior_database` subdirectory of a cloned repository.

```{r}
my_pdb <- pdb_local("/path/to/posteriordb/posterior_database")
# Alternatively: Sys.setenv(PDB_PATH = "/path/to/posterior_database")
```

Local access needs no Stan backend, YAML configuration or GitHub packages.
`pdb_config()` optionally reads `.pdb_config.yml` using `yaml`; its `type`
must be `local` or `github`. GitHub connections require `remotes`, and content
downloads require `httr`. For more details, see `?pdb_local`.

To access the database remotely, use a GitHub connection:

```{r, eval=FALSE}
my_pdb <- pdb_github()
```

When you have a connection to the posterior database of choice, you can access the data, models etc., using the same functionality.

## Choose a workflow

| Task | Entry point |
| --- | --- |
| Read an existing posterior | `posterior(name, my_pdb)`, then data/model/reference getters |
| Build a posterior from in-memory components | `as.posterior(list(pdb_model_code = code, pdb_data = data, dimensions = counts))` |
| Prepare a contribution from a completed RStan fit | `create_pdb_bundle(fit, data = inputs, ...)`, then inspect and explicitly write the bundle |
| Import a completed RStan or CmdStanR fit for an existing posterior | `import_reference_posterior_draws(fit, posterior = name, pdb = my_pdb, write = FALSE)` |
| Run several existing reference workflows sequentially | `sequential_batch_workflow(reference_posteriors, sampling, pdb = my_pdb)` |

The [bundle and fit-import guide](doc/CREATE_PDB_BUNDLE.md) explains metadata,
variable selection, diagnostics, writing, batch argument forms and recovery.
`dimensions` records positive unconstrained parameter counts, rather than saved
array shapes. CmdStanR fit import needs supporting files and enough CSV precision
for constrained values (`sig_figs = 18` when sampling). Bundle construction
currently supports RStan.

`create_pdb_bundle()` builds in memory. With `check = FALSE` it defers diagnostic
calculation; structural validation still applies. The existing-posterior importer
always checks draws and returns failed candidates for inspection when
`write = FALSE`. Accepted reference draws require exactly 10,000 retained draws,
at least four chains and the documented diagnostic gates. ESS is informational.

Explicit bundle/component writes are sequential: an I/O error can leave earlier
components on disk, and the error identifies the failed destination and completed
bundle components. Inspect those files before retrying and review them in the
contribution pull request. Existing-posterior fit import stages and verifies its
files together and attempts rollback on failure. Refresh the connection or clear
its cache after ordinary component writes/removals.

For contributions, follow the [contribution guide](vignettes/contributing.Rmd)
and submit a pull request to PosteriorDB. Neither preparing nor writing a local
bundle publishes it. All examples below are illustrative and are not executed
when rendering this README; available content depends on the selected database.

## Access content

To list the posteriors in the database, use `posterior_names()`.

```{r}
pos <- posterior_names(my_pdb)
head(pos)
```

In the same fashion, we can list data and models included in the database as

```{r}
mn <- model_names(my_pdb)
head(mn)

dn <- data_names(my_pdb)
head(dn)
```

We can also get all information on each posterior as a table with

```{r}
pos <- posteriors_tbl_df(my_pdb)
head(pos)
```

A posterior record links its data and model by metadata. Its saved name need
not be derived from those component names. To access a saved posterior, use
its name:

```{r}
po <- posterior("eight_schools-eight_schools_centered", my_pdb)
```

From the posterior object, we can access data, model code (i.e., Stan code in this case) and other useful information.

```{r}
dat <- pdb_data(po)
dat

code <- stan_code(po)
code
```

We can also access the paths to data after they have been unzipped and copied to the cache directory set in `pdb` (the R temp directory by default).

```{r}
dfp <- data_file_path(po)
dfp

scfp <- stan_code_file_path(po)
scfp
```

We can also access information regarding the model and the data used to compute the posterior.

```{r}
data_info(po)
model_info(po)
```

Citation keys refer to BibTeX items found in `content/references/references.bib`.

We can access most of the posterior information as a `tbl_df` using

```{r}
tbl <- posteriors_tbl_df(my_pdb)
head(tbl)
```

Use `filter_posteriors()` to return matching posterior objects. This optional
workflow requires `dplyr` and follows its filter syntax; listing metadata tables
with `posteriors_tbl_df()` does not require `dplyr`.

```{r}
pos <- filter_posteriors(pdb = my_pdb, data_name == "eight_schools")
pos
```

To access reference posterior draws, we use `reference_posterior_draws()`.

```{r}
rpd <- reference_posterior_draws(po)
```

`reference_posterior_draws()` returns a `pdb_reference_posterior_draws` object
that also inherits from `posterior::draws_list`. The dotted
`as.reference_posterior_draws()` wraps existing draw arrays/lists; the underscore
`as_reference_posterior_draws()` converts and checks a completed Stan fit.
Summarize stored draws with the `posterior` package:

```{r}
posterior::summarize_draws(rpd)
```

To access information on the reference posterior we can use `reference_posterior_draws_info()` or use `info()` on the reference posterior. The posterior reference draws return information on how the reference posterior was computed.

```{r}
rpi <- reference_posterior_draws_info(po)
rpi

info(rpd)
```

## Running tests

The standard suite uses local fixtures. It does not clone PosteriorDB or run
Stan compilation/MCMC or live GitHub requests:

```r
pkgload::load_all(".", helpers = FALSE, export_all = FALSE)
testthat::test_dir("tests/testthat", package = "posteriordb")
```

Opt in separately with `PDB_TEST_DATABASE=true`, `PDB_TEST_STAN=true`, or
`PDB_TEST_GITHUB=true`. Corpus tests also need an existing checkout in
`PDB_PATH` (or `options(pdb_path=...)`, which takes precedence). Each corpus
test copies that checkout and uses a temporary working directory, restoring
settings and removing its copy even on failure. This also applies to
`testthat::test_file()` and `devtools::test()` through the safety helper. Keep
helper loading enabled. Tests never clone a database implicitly.

Real Stan tests require the selected backend and compiler; live GitHub tests
also require the existing credentials. The older contribution integration
test runs a substantial sampling job and needs both database and Stan flags.
Use a chosen PosteriorDB commit when comparing corpus results. The manual
`Database integration` workflow takes a full commit SHA; ordinary check and
coverage jobs run the offline suite. External corpus compatibility and real
backend results are reported separately from fixture tests.
