---
title: "Add Content to the Posterior Database using R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add Content to the Posterior Database using R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

First, clone the `posteriordb` repository for a local copy you want to update. Then install the posteriordb R package.

For a continuous eight-schools example that constructs, checks, writes, and
reads back all linked objects with `create_pdb_bundle()`, see the
[end-to-end bundle workflow](../doc/CREATE_PDB_BUNDLE.md#end-to-end-example-eight-schools).
The steps below describe the individual-object contribution workflow.

Writes update this local checkout; contributing the files requires a commit
and pull request for review. Bundle and individual component writes happen
sequentially. If an I/O or serialization error occurs, the operation stops
and retains files already written or replaced. The failure message names
the destination; bundle failures also list completed components. The failing
component may contain only some of its files, and a failed ZIP command keeps
the uncompressed JSON. A failed call returns an error rather than a success
report. Diagnostic rejection is a separate outcome: a bundle can write its
data, model and posterior while skipping rejected reference draws.

After an I/O failure, inspect `git status`, `git diff`, and any new untracked
files in the database checkout. Fix the cause, review incomplete files, and
rerun the workflow. Choose `overwrite = TRUE` only after reviewing the files
to be replaced; reused bundle components are never overwritten, so resolve
incomplete reused components first. Refresh the connection or clear its
cache before checking changed files. Check the completed metadata, links and
payloads before committing and opening the pull request. Reviewers can then
assess the proposed files. Automatic rollback for these writes remains an
open issue for future discussion; the current workflow keeps partial writes
for inspection. See the [bundle workflow guide](../doc/CREATE_PDB_BUNDLE.md)
for bundle creation, diagnostic results and write reports.

```{r, message=FALSE, eval=FALSE}
remotes::install_github("stan-dev/posteriordb-r")
```

As a first step, we load the `posteriordb` R package and create a connection to the local posterior database. If you intend to add stan-code, also load `rstan`.

```{r, message=FALSE, eval=FALSE}
library(posteriordb)
library(rstan)

# Set the environment variable PDB_PATH to your local posteriordb repo as
Sys.setenv(PDB_PATH = "[path to local posteriordb]")

# We setup a connection to the local posteriordb
pdbl <- pdb_local()

```

```{r, message=FALSE, eval=TRUE, echo=FALSE}
library(posteriordb)
library(rstan)

tmp <- tempdir(check = TRUE)
git2r::clone("https://github.com/stan-dev/posteriordb",
             local_path = file.path(tmp, "posteriordb"))
Sys.setenv(PDB_PATH = file.path(tmp, "posteriordb"))
# We setup a connection to the local posteriordb
pdbl <- pdb_local()
```

Since a posterior consist of data and a model, we start by adding those parts.

## Adding the Data

The next step is to add and write down information about the data you want to add. Below is a test example using data from the eight schools example. Information on the different parts of the metadata can be found in the file [github.com/stan-dev/posteriordb/doc/DATABASE_CONTENT.md](https://github.com/stan-dev/posteriordb/blob/master/doc/DATABASE_CONTENT.md).

```{r}
x <- list(name = "test_eight_schools_data",
          keywords = c("test_data"),
          title = "A Test Data for the Eight Schools Model",
          description = "The data contain data from eight schools on SAT scores.",
          urls = "https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html",
          references = "rubin1981estimation",
          added_date = Sys.Date(),
          added_by = "Stanislaw Ulam")

# Create the data info object
di <- as.pdb_data_info(x)

# Setup the data for eight schools
eight_schools <- list(J = 8L,
                      y = as.integer(c(28, 8, -3, 7, -1, 1, 18, 12)),
                      sigma = as.integer(c(15, 10, 16, 11, 9, 11, 10, 18)))

# Create the data object
dat <- as.pdb_data(eight_schools, info = di)

dat
info(dat)
```


We can now add the data object to the database. The function ```write_pdb()``` will zip and write the data. By setting `overwrite = TRUE`, the function will overwrite the data if it already exists. If `overwrite = FALSE`, the function would throw an error if the file already exists.

```{r, }
write_pdb(dat, pdbl, overwrite = TRUE)
```

If we want to remove the data object, we can use
```{r, eval=FALSE}
remove_pdb(dat, pdbl)
```

Note that we now added a reference `rubin1981estimation` that already exists in the bibliography. If a new reference is added, it also needs to be added to the BibTeX bibliography.



## Adding the Model

Similarly, we can add stan-files and model information as follows (this is a test case using the eight school model). Note that we compile the model code using Stan. Compiling is not necessary to add code but simplify.

Just as with the data, information on the different parts of the metadata can be found in [github.com/stan-dev/posteriordb/doc/DATABASE_CONTENT.md](https://github.com/stan-dev/posteriordb/blob/master/doc/DATABASE_CONTENT.md).

```{r}
x <- list(name = "test_eight_schools_model",
          keywords = c("test_model", "hierarchical"),
          title = "Test Non-Centered Model for Eight Schools",
          description = "An hierarchical model, non-centered parametrisation.",
          urls = c("https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html"),
          framework = "stan",
          references = NULL,
          added_by = "Stanislaw Ulam",
          added_date = Sys.Date())
mi <- as.pdb_model_info(x)

# Read in Stan model and compile the model (using rstan)
smc <- "
data {
  int <lower=0> J; // number of schools
  real y[J]; // estimated treatment
  real<lower=0> sigma[J]; // std of estimated effect
}
parameters {
  vector[J] theta_trans; // transformation of theta
  real mu; // hyper-parameter of mean
  real<lower=0> tau; // hyper-parameter of sd
}
transformed parameters{
  vector[J] theta;
  // original theta
  theta=theta_trans*tau+mu;
}
model {
  theta_trans ~ normal (0,1);
  y ~ normal(theta , sigma);
  mu ~ normal(0, 5); // a non-informative prior
  tau ~ cauchy(0, 5);
}
"
mc <- as.model_code(smc, info = mi, framework = "stan")

# The model object to include
mc
info(mc)
```

Ideally, check that the stan code compiles before it is added to the posterior database. A simple way is to compile the model with `rstan` and directly create the model code object from the stan object.

```{r, eval=FALSE}
sm <- rstan::stan_model(model_code = smc)
mc <- as.model_code(sm, info = mi)
```

To add the model to the local posterior database, we again use:
```{r, eval=FALSE}
write_pdb(mc, pdbl)
```

```{r, echo=FALSE}
write_pdb(mc, pdbl, overwrite = TRUE)
```

And to remove the model object, we use
```{r, eval=FALSE}
remove_pdb(mc, pdbl)
```


## Adding the Posterior

When we have added the data and the model, we can add a posterior object connecting the model and the data. We must add the model code and data before adding the posterior.

Again, information on the different parts of the metadata can be found in [github.com/stan-dev/posteriordb/doc/DATABASE_CONTENT.md](https://github.com/stan-dev/posteriordb/blob/master/doc/DATABASE_CONTENT.md).

The posterior dimensions list records unconstrained parameter counts keyed
by parameter name, not the shapes of saved draws. Transformed parameters and
generated quantities do not contribute independent coordinates. Choose optional
saved outputs for reference-draw diagnostics with `include` and `exclude` when
calling `compute_reference_posterior_draws()`, as shown below.


```{r}
x <- list(pdb_model_code = mc,
          pdb_data = dat,
          keywords = "posterior_keywords",
          urls = "posterior_urls",
          references = "rubin1981estimation",
          dimensions = list("dimensions" = 2, "dim" = 3),
          reference_posterior_name = NULL,
          added_date = Sys.Date(),
          added_by = "Stanislaw Ulam")
po <- as.pdb_posterior(x)
```

As with the data and model object, we then use `write_pdb()` and `remove_pdb()` to add and remove the object from the local posterior database.

```{r, eval = FALSE}
write_pdb(po, pdbl)
```
```{r, echo=FALSE}
write_pdb(po, pdbl, overwrite = TRUE)
```
```{r, eval=FALSE}
remove_pdb(mc, pdbl)
```


## Checking the final posterior, data and model

Finally, we want to check that everything is in order with the posterior. We do this as follows. Here we skip checking that the code compiles.

```{r, eval=TRUE}
check_pdb_posterior(po, run_stan_code_checks = FALSE)
```

If the posterior, model, and data pass all checks, we can add it to the posteriordb. Commit the new files in the database and open a Pull Request with the proposed posterior.

## Add Posterior Reference Draws

If possible, we would like to supply posterior reference draws, i.e., draws of excellent quality from the posterior. The [github.com/stan-dev/posteriordb/doc/REFERENCE_POSTERIOR_DEFINITION.md](https://github.com/stan-dev/posteriordb/blob/master/doc/REFERENCE_POSTERIOR_DEFINITION.md) contain details on quality criteria for reference posteriors.
Information on the different parts of the metadata can be found in [github.com/stan-dev/posteriordb/doc/DATABASE_CONTENT.md](https://github.com/stan-dev/posteriordb/blob/master/doc/DATABASE_CONTENT.md).

```{r}
pdbl <- pdb_local()
po <- posterior("test_eight_schools_data-test_eight_schools_model", pdbl)

# Setup reference posterior info ----
x <- list(name = posterior_name(po),
          inference = list(method = "stan_sampling",
                           method_arguments = list(chains = 10,
                                                   iter = 20000,
                                                   warmup = 10000,
                                                   thin = 10,
                                                   seed = 4711,
                                                   control = list(adapt_delta = 0.92))),
          diagnostics = NULL,
          checks_made = NULL,
          comments = "This is a test reference posterior",
          added_by = "Stanislaw Ulam",
          added_date = Sys.Date(),
          versions = NULL)

# Create a reference posterior draws info object
rpi <- as.pdb_reference_posterior_info(x)
```

The reference posterior draws info contains all information to compute the posterior using rstan. We then check that the reference posterior criteria are fulfilled and add checked diagnostics to the object.

```{r, eval=FALSE}
# Compute the reference posterior
rp <- compute_reference_posterior_draws(rpi, pdbl)
# Check that the draws are of sufficient quality to use as a reference posterior
rp <- check_reference_posterior_draws(x = rp)
```

The default `include = NULL` retains and diagnoses all saved model outputs
except `lp__`, including transformed parameters and generated quantities.
For example, a saved `vector[10] exp_mu` is retained with all ten elements.
To retain everything except that optional output, specify only `exclude`:

```{r, eval=FALSE}
rp <- compute_reference_posterior_draws(rpi, pdbl, exclude = "exp_mu")
rp <- check_reference_posterior_draws(rp)
```

Use `include = "all"` as an alias for the default, or `include = "none"` /
`character(0)` for required parameter-block and dimension-listed variables
only. `exclude = NULL`, `character(0)`, or `"none"` excludes nothing;
`exclude = "all"` keeps required variables only. Explicitly excluding required
variables, or naming the same variable in both selectors, raises an error.
Every retained variable receives diagnostics; a failing retained output
prevents acceptance. Omitted outputs are not certified, and selection does
not change unconstrained counts.

These aliases are consistent across bundle, compute, import, diagnostics,
and count helpers. Standalone diagnostic/count helpers have no mandatory
selection, so `include = "none"` or `exclude = "all"` raises an empty-selection
error. In earlier versions of the package, `compute_reference_posterior_draws()`
kept only the variables listed in the posterior's `dimensions`; use
`include = "none"` to keep only the parameters and those variables. See
[Choose variables](../doc/CREATE_PDB_BUNDLE.md#choose-variables) for examples.

We can now write the reference posterior draws to the posteriordb, just as the other objects.

```{r, eval = FALSE}
write_pdb(rp, pdbl, overwrite = TRUE)
```

We can again check the posterior, and its reference draws as follows.

```{r, eval = FALSE}
check_pdb_posterior(po, pdbl)
```

We can also remove the reference posterior object with
```{r, eval=FALSE}
remove_pdb(rp, pdbl)
```

```{r cleanup, eval=TRUE, echo=FALSE}
# Cleanup
# mc <- pdb_model_code("test_eight_schools_model", pdbl)
invisible(remove_pdb(mc, pdbl))
# dat <- pdb_data("test_eight_schools_data", pdbl)
invisible(remove_pdb(dat, pdbl))
# po <- pdb_posterior("test_eight_schools_data-test_eight_schools_model", pdbl)
invisible(remove_pdb(po, pdbl))
```
