---
title: "Build a reference-draw bundle from a sampled fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a reference-draw bundle from a sampled fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This guide starts with an already sampled `rstan::stanfit`, the exact Stan
input list used for that fit, and human labels for the data and model. Bundle
construction does not compile or sample the model, access a database, or write
files.

## Construct the bundle

Pass data and the two required names and titles. Parameter selection uses base
names, so including an array parameter retains all its saved scalar elements.
The default also includes saved transformed parameters and generated
quantities, while excluding `lp__`.
Use exact, named arguments after `data`; misspellings, duplicate arguments,
and unnamed metadata are rejected.

```{r, eval=FALSE}
bundle <- create_pdb_bundle(
  fit,
  data = stan_data,
  data_info = list(name = "experiment-1", title = "Experiment 1 inputs"),
  model_info = list(name = "normal-model", title = "Normal model"),
  posterior_info = list(added_by = "Contributor name"),
  include = c("mu", "sigma")
)

bundle$data
bundle$model_code
bundle$posterior
bundle$reference_draws
bundle$diagnostics
bundle$provenance

info(bundle$data)            # data metadata, including its derived file path
info(bundle$model_code)      # model metadata and implementation path
info(bundle$reference_draws) # inference metadata and diagnostic evidence
```

`posterior_info$added_by` and `$added_date` supply common defaults. Each of
`data_info`, `model_info`, and `reference_info` can override them for that
object. The constructor derives posterior names, paths, framework, and
dimensions; conflicting structural metadata is rejected. `reference_info`
accepts human comments and contributor information, never diagnostic values
or acceptance flags.

The returned `pdb_reference_bundle` contains the data, model code, posterior,
reference draws, diagnostic report, and provenance. Its posterior embeds the
content needed by `get_data()`, `model_code()`, and
`reference_posterior_draws()`, so those getters work before persistence. A
`pdb` connection may be attached for later use, but it is not read during
construction. File-path methods report that embedded objects have not been
persisted.

## Inspect diagnostics

`create_pdb_bundle()` constructs the in-memory bundle. By default,
`check = TRUE` computes the reference-draw diagnostic report and records
whether the candidate meets the package's acceptance checks. A failed
candidate is returned with its failures; the function does not resample, thin,
or select a target draw count. Use `check = FALSE` to skip diagnostic and
acceptance-metric calculations entirely. In that case
`bundle$diagnostics` is `NULL`, and the reference draws have no acceptance
flags. The separate `reference_draw_diagnostics()` function calculates a
report, while `passes_reference_draw_checks()` returns a logical result for
selected checks. Writing `bundle$reference_draws` with `write_pdb()` requires
all recorded acceptance flags to be `TRUE`.

```{r, eval=FALSE}
reference_draw_diagnostics(fit, checks = "mean_lag1_ac")
passes_reference_draw_checks(fit) # defaults to the lag-1 check only
passes_reference_draw_checks(fit, checks = c("mean_lag1_ac", "r_hat"))
```

The selective calls need only the fit. A lag-only request reads post-warmup
draws and does not calculate ESS or R-hat or read sampler energy diagnostics.
The all-check report includes the observed metrics, policy thresholds, each
check's status, and variables or chains that failed.
Undefined metrics, including autocorrelation for constant quantities, fail
their selected checks while the rest of the report remains available.

Full acceptance requires exactly 10,000 retained draws across at least four
chains, mean absolute per-chain lag-1 autocorrelation at most 0.05 for every
variable, R-hat at most 1.01, E-FMI at least 0.2 for each chain, and zero
divergences. ESS and treedepth are informational. Passing a selected check
alone does not establish full reference acceptance.

## Data provenance and limits

Supply the exact named Stan input list after preprocessing. Explicit `list()`
declares that the model had no data inputs; omitted or `NULL` data currently
errors because automatic recovery is not implemented. The package records
whether data was caller-supplied but cannot prove that it produced the fit.
Structural checks preserve the supplied R values and dimensions but cannot
establish scientific equivalence.
Data must contain ordinary finite numeric, integer, or logical Stan inputs,
including vectors and arrays; preprocess raw data frames before calling.

Saved array elements are checked against the fit's declared dimensions,
including singleton and non-square axes. Partially saved variables cannot be
included. Zero-sized selected variables and source code with unresolved
`#include` directives produce clear errors. The first bundle backend supports
completed RStan HMC fits; other inference algorithms are rejected. Current
package versions describe the import operation in `provenance$import_versions`,
separately from fit-derived `provenance$sampling_metadata`. An unavailable
original sampling version is left unknown.

Pass one completed fit. The constructor does not combine sampling runs or
merge database objects. `fit_timestamp`, when available, is the fit's recorded
date, which is not necessarily its original sampling date.

## Explicit persistence

Creating a bundle creates R objects and their metadata in memory, even if
`pdb` is supplied or those names already exist in the database. It does not
create, merge, or replace files. Attaching `pdb` does not authorize writes.

If you choose to persist new objects, use the existing writers explicitly:

```{r, eval=FALSE}
write_pdb(bundle$data, pdb = my_pdb, overwrite = FALSE)
write_pdb(bundle$model_code, pdb = my_pdb, overwrite = FALSE)
write_pdb(bundle$posterior, pdb = my_pdb, overwrite = FALSE)
# Only checked, passing reference draws may be written:
write_pdb(bundle$reference_draws, pdb = my_pdb, overwrite = FALSE)
```

The data and model writers also write their respective info files. The
posterior writer writes its metadata and links, without the embedded payloads.
These separate writes are not an atomic all-object import. Review the target
names first and reload persisted posteriors to obtain database-backed paths.

Existing destination files are protected by the default `overwrite = FALSE`.
Use new names for new contributions; only use `overwrite = TRUE` when you
intend and are authorized to replace those specific existing objects.
Contributor metadata such as `added_by` records attribution; the existing
writers do not implement an ownership or access-control system.
