---
title: "Build a reference-draw bundle from a sampled fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a reference-draw bundle from a sampled fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This guide starts with an already sampled `rstan::stanfit`, the exact Stan
input list used for that fit, and human labels for the data and model. Bundle
construction does not compile or sample the model, access a database, or write
files.

## Construct the bundle

Pass data and the two required names and titles. Parameter selection uses base
names, so including an array parameter retains all its saved scalar elements.
The default also includes saved transformed parameters and generated
quantities, while excluding `lp__`.

```{r, eval=FALSE}
bundle <- create_pdb_reference_draws(
  fit,
  data = stan_data,
  data_info = list(name = "experiment-1", title = "Experiment 1 inputs"),
  model_info = list(name = "normal-model", title = "Normal model"),
  posterior_info = list(added_by = "Contributor name"),
  include = c("mu", "sigma")
)

bundle$data
bundle$model_code
bundle$posterior
bundle$reference_draws
bundle$diagnostics
bundle$provenance
```

The returned `pdb_reference_bundle` contains the data, model code, posterior,
reference draws, diagnostic report, and provenance. Its posterior embeds the
content needed by `get_data()`, `model_code()`, and
`reference_posterior_draws()`, so those getters work before persistence. A
`pdb` connection may be attached for later use, but it is not read during
construction. File-path methods report that embedded objects have not been
persisted.

## Inspect diagnostics

The fit may be useful even when it does not meet every reference-draw
criterion. Failed checks are reported, and the candidate is returned without
resampling, thinning, or selecting a target draw count. Use `check = FALSE` to
leave the candidate explicitly unchecked; this does not set acceptance flags.
Writers continue to require a checked, passing reference draw object.

```{r, eval=FALSE}
reference_draw_diagnostics(fit, checks = "mean_lag1_ac")
passes_reference_draw_checks(fit, checks = c("mean_lag1_ac", "r_hat"))
```

The selective calls need only the fit. A lag-only request reads post-warmup
draws and does not calculate ESS or R-hat or read sampler energy diagnostics.
The all-check report includes the observed metrics, policy thresholds, each
check's status, and variables or chains that failed.

## Data provenance and limits

Supply the exact named Stan input list after preprocessing. Explicit `list()`
declares that the model had no data inputs; omitted or `NULL` data currently
errors because automatic recovery is not implemented. The package records
whether data was caller-supplied but cannot prove that it produced the fit.
Structural checks preserve the supplied R values and dimensions but cannot
establish scientific equivalence.

This bundle is an in-memory construction API. It does not perform an atomic
database import of all contained objects. Persist objects using the package's
existing writers after review, and reload persisted posteriors to obtain their
database-backed file paths.
