---
title: "Build a reference-draw bundle from a sampled fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a reference-draw bundle from a sampled fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This guide starts with an already sampled `rstan::stanfit`, the exact Stan
input list used for that fit, and a name and title for the data and the model.
Building a bundle does not sample the model or write any files.

It is a short introduction. The
[bundle guide](../doc/CREATE_PDB_BUNDLE.md) in the package repository has a
complete worked example and covers reusing existing database records,
choosing variables, and importing CmdStanR fits.

## Construct the bundle

Pass data and the two required names and titles. Parameter selection uses base
names, so including an array parameter retains all its saved scalar elements.
The default also includes saved transformed parameters and generated
quantities, while excluding `lp__`.
Use exact, named arguments after `data`; misspellings, duplicate arguments,
and unnamed metadata are rejected.

```{r, eval=FALSE}
bundle <- create_pdb_bundle(
  fit,
  data = stan_data,
  data_info = list(name = "experiment_1", title = "Experiment 1 inputs"),
  model_info = list(name = "normal_model", title = "Normal model"),
  posterior_info = list(added_by = "Contributor name"),
  include = c("mu", "sigma")
)

bundle$data
bundle$model_code
bundle$posterior
bundle$reference_draws
bundle$diagnostics
bundle$provenance

info(bundle$data)            # data metadata, including its derived file path
info(bundle$model_code)      # model metadata and implementation path
info(bundle$reference_draws) # inference metadata and diagnostic evidence
```

`posterior_info$added_by` and `$added_date` supply common defaults. Each of
`data_info`, `model_info`, and `reference_info` can override them for that
object. The constructor derives posterior names, paths, framework, and
dimensions; conflicting structural metadata is rejected. `reference_info`
accepts human comments and contributor information, never diagnostic values
or acceptance flags.

The returned `pdb_reference_bundle` contains the data, model code, posterior,
reference draws, diagnostic report, and provenance. Its posterior embeds the
content needed by `get_data()`, `model_code()`, and
`reference_posterior_draws()`, so those getters work before anything is saved.
If you supply `pdb`, the connection is attached to the new objects, and it is
used to look up any data, model or posterior you pass by its saved name.

## Inspect diagnostics

`create_pdb_bundle()` constructs the in-memory bundle. By default,
`check = TRUE` computes the reference-draw diagnostic report and records
whether the candidate meets the package's acceptance checks. A failed
candidate is returned with its failures; the function does not resample, thin,
or select a target draw count. Use `check = FALSE` to skip diagnostic and
acceptance-metric calculations entirely. In that case
`bundle$diagnostics` is `NULL`, and the reference draws have no acceptance
flags. The separate `reference_draw_diagnostics()` function calculates a
report, while `passes_reference_draw_checks()` returns a logical result for
selected checks. Writing `bundle$reference_draws` with `write_pdb()` requires
all recorded acceptance flags to be `TRUE`.

To defer the full check, pass the unchecked bundle to
`check_reference_posterior_draws()`. It returns an updated bundle with the
report attached and acceptance flags set only if every required check passes.
Failed candidates keep their failure report and remain unwritable.

```{r, eval=FALSE}
bundle <- create_pdb_bundle(fit, data = stan_data,
  data_info = data_info, model_info = model_info, check = FALSE)
bundle <- check_reference_posterior_draws(bundle)
bundle$diagnostics$status
# If every required check passed:
write_pdb(bundle$reference_draws, target_pdb)
```

```{r, eval=FALSE}
reference_draw_diagnostics(fit, checks = "mean_lag1_ac")
passes_reference_draw_checks(fit) # every requirement
passes_reference_draw_checks(fit, checks = c("mean_lag1_ac", "r_hat"))
```

The selective calls need only the fit. A lag-only request reads post-warmup
draws and does not calculate ESS or R-hat or read sampler energy diagnostics.
The all-check report includes the observed metrics, policy thresholds, each
check's status, and variables or chains that failed.
Undefined metrics, including autocorrelation for constant quantities, fail
their selected checks while the rest of the report remains available.

Full acceptance requires exactly 10,000 retained draws across at least four
chains, mean absolute per-chain lag-1 autocorrelation at most 0.05 for every
variable, R-hat at most 1.01, E-FMI at least 0.2 for each chain, and zero
divergences. ESS and treedepth are informational. Passing a selected check
alone does not establish full reference acceptance.

## Data provenance and limits

Supply the exact named Stan input list after preprocessing. Use `list()` for
a model with no data. `data` is required: it cannot be recovered from the fit,
and the package cannot confirm that the list you supply produced the fit.
Data must contain ordinary finite numeric, integer, or logical Stan inputs,
including vectors and arrays; preprocess raw data frames before calling.

Saved array elements are checked against the fit's declared dimensions,
including singleton and non-square axes. Partially saved variables cannot be
included. Zero-sized selected variables and source code with unresolved
`#include` directives produce errors. Bundles are built from completed RStan
HMC fits; other inference algorithms are rejected.
`provenance$import_versions` records the package versions used to build the
bundle, and `provenance$sampling_metadata` records what the fit itself reports
about sampling. Neither proves which versions were used for sampling.

Pass one completed fit. The constructor does not combine sampling runs or
merge database objects. `fit_timestamp`, when available, is the fit's recorded
date, which is not necessarily its original sampling date.

## Write to a database

Creating a bundle only creates R objects, even if `pdb` is supplied or those
names already exist in the database. To save it, write the whole bundle:

```{r, eval=FALSE}
result <- write_pdb(bundle, my_pdb, overwrite = FALSE)
result$written
result$reference_draws_written
```

New data, model and posterior records are written. Reference draws and their
summary statistics are written only if every check passed; if a check failed,
the other records are still written and `result$skipped_reason` says why.
Records reused from the database are left as they are.

You can also write the objects one at a time:

```{r, eval=FALSE}
write_pdb(bundle$data, pdb = my_pdb, overwrite = FALSE)
write_pdb(bundle$model_code, pdb = my_pdb, overwrite = FALSE)
write_pdb(bundle$posterior, pdb = my_pdb, overwrite = FALSE)
# Only checked, passing reference draws may be written:
write_pdb(bundle$reference_draws, pdb = my_pdb, overwrite = FALSE)
```

Each write is separate, so if a later one fails the earlier files remain.

With the default `overwrite = FALSE`, an existing file with the same name is
an error. Use new names for new contributions, and set `overwrite = TRUE` only
when you mean to replace those existing files.
