---
output:
  rmarkdown::md_document:
    variant: gfm
    toc: true
editor_options:
  markdown:
    wrap: 72
---

<!-- CREATE_PDB_BUNDLE.md is generated from CREATE_PDB_BUNDLE.Rmd. Please edit that file. -->

# Create and contribute a reference-draw bundle

A PosteriorDB contribution has four linked parts: the **data**, the
**model code**, the **posterior** that names both, and the **reference
draws** for that posterior. `create_pdb_bundle()` builds all four from a
Stan fit you have already sampled, checks the draws, and returns them as
one R list. Nothing is written until you call `write_pdb()`.

``` r
bundle <- create_pdb_bundle(fit, data = stan_data,
  data_info = list(name = "my_data", title = "My data"),
  model_info = list(name = "my_model", title = "My model", framework = "stan"))
bundle$diagnostics$status          # did the draws pass?
write_pdb(bundle, pdb_local("/path/to/posteriordb/posterior_database"))
```

`create_pdb_bundle()` accepts RStan `stanfit` objects. If you sampled
with CmdStanR, or only want to add draws to a posterior that is already
in the database, use `import_reference_posterior_draws()` instead; see
[Import a fit for an existing
posterior](#import-a-fit-for-an-existing-posterior).

## Which function do I need?

| You have | You want | Use |
|------------------------|------------------------|------------------------|
| A `stanfit`, its data, and nothing in the database yet | Data, model, posterior and draws | `create_pdb_bundle()` then `write_pdb()` |
| A `stanfit` and some records already in the database | The missing records and draws | `create_pdb_bundle()` with existing objects or names |
| A `stanfit` or CmdStanR fit, and the posterior already exists | Reference draws only | `import_reference_posterior_draws()` |
| A fit, and you only want to look at diagnostics | A report, nothing saved | `reference_draw_diagnostics()` |

Two similarly named functions do different jobs:

- `as_reference_posterior_draws(fit, posterior, pdb)` (underscores)
  converts a completed fit and runs the acceptance checks.
- `as.reference_posterior_draws(draws, info)` (dots) only attaches
  metadata to draws you prepared yourself. It does not check them.

## End-to-end example: eight schools

This example samples the non-centered eight-schools model, builds a
bundle, checks it, writes it to a local database checkout, and reads it
back. You need RStan and a working C++ toolchain. Replace the database
path and contributor name, and choose data and model names that are not
already used in your database.

The citation key `rubin1981estimation` must already be in the database
bibliography; see [Add BibTeX references](#add-bibtex-references) if you
need to add one.

```{r eight-schools-end-to-end, eval=FALSE}
library(posteriordb)
library(rstan)

# 1. Connect to your local PosteriorDB checkout.
database_path <- "/path/to/posteriordb/posterior_database"
pdbl <- pdb_local(database_path)
stopifnot("rubin1981estimation" %in% bibliography_keys(pdbl))

# 2. Sample, keeping the exact input list used for sampling.
source(system.file("test_files/eight_schools.R", package = "posteriordb"))
stan_file <- system.file(
  "test_files/eight_schools_noncentered.stan", package = "posteriordb"
)
stan_model <- rstan::stan_model(file = stan_file, auto_write = FALSE)
fit <- rstan::sampling(
  stan_model, data = eight_schools,
  chains = 10, iter = 20000, warmup = 10000, thin = 10,
  seed = 4711, control = list(adapt_delta = 0.99), refresh = 0
)
# 1,000 retained draws per chain: 10,000 draws in total.

# 3. Build the linked objects in memory and check the draws.
bundle <- create_pdb_bundle(
  fit, data = eight_schools, added_by = "Your Name",
  data_info = list(
    name = "test_eight_schools_data",
    title = "Eight schools treatment estimates",
    description = "Treatment estimates and standard errors for eight schools.",
    references = "rubin1981estimation"
  ),
  model_info = list(
    name = "test_eight_schools_model",
    title = "Non-centered eight schools model",
    framework = "stan",
    description = "A hierarchical model with a non-centered parameterization.",
    references = "rubin1981estimation"
  ),
  posterior_info = list(references = "rubin1981estimation"),
  reference_info = list(comments = "Eight schools bundle contribution."),
  pdb = pdbl
)
bundle$posterior$name
bundle$posterior$dimensions
# theta_trans = 8, mu = 1, tau = 1

# 4. Inspect the checks before writing anything.
bundle$diagnostics$status
bundle$diagnostics$failures
assert_checked_reference_posterior_draws(bundle$reference_draws)

# 5. Write the bundle and look at what was written.
write_result <- write_pdb(bundle, pdbl, overwrite = FALSE)
write_result[c("written", "reused", "reference_draws_written",
               "summary_statistics_written", "skipped_reason")]
bundle <- write_result$bundle

# 6. Read the saved contribution back.
read_cache <- tempfile("readback-")
dir.create(read_cache)
read_pdb <- pdb_local(database_path, cache_path = read_cache)
saved_posterior <- posterior(bundle$posterior$name, pdb = read_pdb)
check_pdb_posterior(saved_posterior, run_stan_code_checks = FALSE)
saved_draws <- reference_posterior_draws(saved_posterior)
posterior::variables(posterior::as_draws_array(saved_draws))
# theta_trans[1:8], mu, tau and theta[1:8]
```

The assertion in step 4 stops the script if any check failed. These
sampling settings give the required number of draws, but no settings can
guarantee that every check passes; see [When a check
fails](#when-a-check-fails).

Finally, review the changes in your database checkout, commit them
there, and open a pull request against PosteriorDB. The R package and the
PosteriorDB database are separate repositories; the contribution files
belong in the database.

``` sh
git -C /path/to/posteriordb status --short
git -C /path/to/posteriordb diff
```

`git diff` does not show the contents of new, untracked files, so open
the new metadata and Stan source as well.

## What is in a bundle

```{r, eval=FALSE}
bundle$data
bundle$model_code
bundle$posterior
bundle$reference_draws
bundle$summary_statistics$mean_value
bundle$summary_statistics$mean_squared_value
bundle$diagnostics
bundle$provenance
```

The posterior carries its data, model code and reference draws with it,
so `get_data()`, `model_code()` and `reference_posterior_draws()` work
on `bundle$posterior` before anything is saved. `bundle$provenance`
records the fit class, the selected variables and the sampling settings
read from the fit. `bundle$summary_statistics` is filled in only when
the draws pass their checks; otherwise it is `NULL`.

> **`dimensions` are parameter counts, not shapes.** In
> `bundle$posterior$dimensions`, each entry is the number of
> *unconstrained* parameters for one model parameter. A `matrix[2, 3] B`
> has `B = 6`, and a `simplex[3]` has 2, even though three values are
> saved for it. Transformed parameters and generated quantities are
> saved as draws but have no entry in `dimensions`. In the eight-schools
> example there are 18 saved columns and 10 unconstrained parameters.

## Supply metadata

`data_info` and `model_info` each need a `name` and a `title`. Names
become file names, so they must be a single path component: dots and
hyphens are fine, slashes are not.

| Argument | Required fields | Optional fields |
|------------------------|------------------------|------------------------|
| `data_info` | `name`, `title` | `description`, `references`, `urls`, `keywords` |
| `model_info` | `name`, `title` | `description`, `references`, `urls`, `keywords`, `prior`, `licence`, `framework` or `model_implementations` |
| `posterior_info` | none | `references`, `urls`, `keywords` |
| `reference_info` | none | `comments` |

`added_by` and `added_date` apply to all four objects. `added_by`
defaults to your system user name and `added_date` to today. Any of the
four lists can carry its own `added_by` or `added_date` to override the
shared value for that object.

Only `fit`, `data`, `added_by` and `added_date` can be passed by
position. Pass everything else by name; unknown or misspelled arguments
and unsupported metadata fields are errors.

For the model, `framework = "stan"` is enough. The model file is stored
as `models/stan/<name>.stan` with the Stan version requirement
`">=2.26.0"`. To record a different requirement, supply the
implementation yourself:

``` r
model_info = list(
  name = "normal",
  title = "Normal model",
  model_implementations = list(stan = list(
    model_code = "models/stan/normal.stan",
    stan_version = ">=2.35.0"
  ))
)
```

`references` are citation keys from the database bibliography, not
BibTeX text. `model_info$prior` is optional descriptive metadata, for
example `prior = list(keywords = "stan_recommended_35dbfe6")`; priors
are not inferred from the Stan code.

The posterior name, its links to the data and model, and its
`dimensions` are worked out for you. You may supply them in
`posterior_info`, but only values that match are accepted.

### Data

Pass the exact named list you gave to Stan, after any preprocessing. Use
`list()` for a model with no data. The package records the list you
supply; it cannot confirm that this was the data used for the fit, and
it cannot recover data from the fit.

## Reuse existing records

If the data, model or posterior is already in the database, pass it
instead of describing it again, either as an object or by its saved
name:

```{r, eval=FALSE}
pdbl <- pdb_local("/path/to/posteriordb/posterior_database")
bundle <- create_pdb_bundle(
  fit,
  data = "existing_data",
  model_code = "existing_model",
  pdb = pdbl
)
```

- If you pass only `posterior`, its linked data and model code are
  looked up for you.
- Leave out the `*_info` list for anything you reuse. If you supply
  both, you get a warning and the existing record's metadata is used.
- An existing model's Stan source must match the source in the fit, and
  an existing posterior's `dimensions` must match the counts inferred
  from the fit. Posteriors that store output shapes in `dimensions`
  rather than unconstrained counts fail this check.

When you write the bundle, reused records are left exactly as they are;
see [Write the bundle](#write-the-bundle).

## Check the draws

`check = TRUE` is the default. The checks run during construction and
the report is stored in the bundle:

```{r, eval=FALSE}
bundle$diagnostics$metrics   # the measured values
bundle$diagnostics$status    # one TRUE/FALSE per check
bundle$diagnostics$failures  # which variables or chains failed
```

Reference draws are accepted when all of these hold:

| Check | Requirement |
|------------------------|------------------------------------------------|
| Number of draws | Exactly 10,000 retained draws |
| Number of chains | At least 4 |
| Lag-1 autocorrelation | Mean absolute value across chains at most 0.05, for every retained variable |
| R-hat | At most 1.01, for every retained variable |
| E-FMI | At least 0.2, for every chain |
| Divergent transitions | None |

Bulk and tail ESS and tree depth are reported but do not decide
acceptance.

A failed check does not raise an error. You still get the bundle back,
with the failures recorded, so that you can inspect them. Draws that
failed, or that have not been checked, cannot be written.

To build the objects first and check later, use `check = FALSE` and then
`check_reference_posterior_draws()`. The sampler diagnostics are kept in
the bundle, so nothing is resampled:

```{r, eval=FALSE}
unchecked <- create_pdb_bundle(
  fit,
  data = eight_schools,
  data_info = list(name = "test_eight_schools_data", title = "Eight schools data"),
  model_info = list(name = "test_eight_schools_model", title = "Eight schools model"),
  check = FALSE
)
is.null(unchecked$diagnostics)  # TRUE

bundle <- check_reference_posterior_draws(unchecked)
bundle$diagnostics$status
```

A bundle is always checked against the full set of requirements. To look
at a single check on a fit, for example while tuning sampler settings,
use the standalone functions. They work on RStan and CmdStanR fits:

```{r, eval=FALSE}
report <- reference_draw_diagnostics(fit, checks = "mean_lag1_ac")
report$metrics
report$thresholds
report$failures
passes_reference_draw_checks(fit, checks = c("mean_lag1_ac", "r_hat"))
```

Passing a selected check does not make draws writable; only the full
check does.

### When a check fails

Fix the sampling, then build a new bundle from the new fit.

| Failed check | What usually helps |
|------------------------|------------------------------------------------|
| Number of draws | Choose `chains`, `iter`, `warmup` and `thin` so that exactly 10,000 draws are retained |
| Lag-1 autocorrelation | Thin more, with correspondingly more iterations |
| R-hat | Longer warmup or more iterations; check for multimodality |
| E-FMI | Reparameterize the model |
| Divergent transitions | Raise `adapt_delta`, or reparameterize |

Every retained variable is checked, including transformed parameters and
generated quantities. A constant output has no defined R-hat or
autocorrelation and therefore fails; exclude it as described below.

## Choose variables

By default a bundle keeps every saved model output except `lp__`:
parameters, transformed parameters and generated quantities. Use
`include` and `exclude` to change that. Both take base names, so
`include = "theta"` selects all eight saved elements of `theta`.

Parameters (the variables in the Stan `parameters` block) are always
kept, because they are needed to reproduce the other outputs. Excluding
one is an error.

| Value | In `include` | In `exclude` |
|-------------------|---------------------------|---------------------------|
| `NULL` (default) | Keep everything | Remove nothing |
| `"all"` | Keep everything | Keep only the parameters |
| `"none"` or `character(0)` | Keep only the parameters | Remove nothing |
| Names | Keep these, plus the parameters | Remove these |

```{r, eval=FALSE}
# Everything except one derived output:
create_pdb_bundle(fit, data = stan_data, data_info = data_info,
                  model_info = model_info, exclude = "exp_mu")

# Parameters only:
create_pdb_bundle(fit, data = stan_data, data_info = data_info,
                  model_info = model_info, include = "none")
```

Things to watch for:

- `c()` is `NULL` in R, so `include = c()` keeps everything. Use
  `"none"` or `character(0)` to keep only the parameters.
- Naming the same variable in both `include` and `exclude` is an error.
- Unknown, duplicated or empty names are errors. `lp__` is never kept.
- Checks and summary statistics cover what you keep. Acceptance says
  nothing about outputs you left out.
- Selection does not change `dimensions`, which always describe the
  model's parameters.

The same `include` and `exclude` rules are used by
`import_reference_posterior_draws()`, `as_reference_posterior_draws()`,
`compute_reference_posterior_draws()`, `reference_draw_diagnostics()`
and `infer_posterior_dimensions()`. For import and compute, variables
listed in the posterior's `dimensions` are also always kept. The
diagnostic and dimension functions have no mandatory variables, so a
selection that leaves nothing is an error there.

Earlier versions of the package kept only the required variables when
computing or importing draws. They now keep every saved output, so an
existing script may return more variables, and may fail a check because
of one of them. Use `include = "none"` to get the earlier selection.

After construction you can still narrow or thin reference draws with
`subset(draws, variable = ...)` and `posterior::thin_draws()`. Both
clear the recorded check results, so check the draws again before
writing them.

## Write the bundle

One call writes the whole bundle:

```{r, eval=FALSE}
write_result <- write_pdb(bundle, pdbl, overwrite = FALSE)
write_result$written
write_result$reused
write_result$reference_draws_written
write_result$summary_statistics_written
write_result$skipped_reason
bundle <- write_result$bundle
```

What gets written:

- **New data, model and posterior records** are written.
- **Records you reused from this database** are left in place and
  listed in `write_result$reused`. They are never overwritten, whatever
  `overwrite` is set to. If a reused posterior has no reference draws
  linked yet, the link is filled in once the draws are written. If it is
  already linked to different reference draws, the write is refused.
- **Reference draws and the two summary statistics** (`mean_value` and
  `mean_squared_value`) are written only if every check passed. If the
  bundle has not been checked yet, it is checked first.

> **A failed check does not stop the write.** The data, model and
> posterior are still written, and only the draws and summaries are
> skipped. `write_result$reference_draws_written` is then `FALSE` and
> `write_result$skipped_reason` says why. If you want all or nothing,
> call `assert_checked_reference_posterior_draws(bundle$reference_draws)`
> first, as in the end-to-end example.

Before anything is written, every destination file is checked:

- With `overwrite = FALSE` (the default), an existing file for a new
  record stops the whole write, and the error lists every file in the
  way. Nothing is written.
- With `overwrite = TRUE`, those files are replaced and listed in
  `write_result$collisions` and `write_result$overwritten`. This
  replaces whatever is stored under those names; it does not merge
  metadata.
- A record reused from a *different* database is copied in if the
  destination has no files with that name. If it does, the write is
  refused under both settings.

The files are written one after another: data, model, posterior, then
draws and summaries. If a disk or serialization error interrupts the
write, the files already written stay in place and the error message
lists the completed parts. Check `git status` in your database checkout,
fix the cause, and write again; files left from the first attempt need
`overwrite = TRUE`.

### Write the objects one at a time

You can also write each object yourself. The posterior must be saved
before its reference draws:

```{r, eval=FALSE}
assert_checked_reference_posterior_draws(bundle$reference_draws)
write_pdb(bundle$data, pdbl, overwrite = FALSE)
write_pdb(bundle$model_code, pdbl, overwrite = FALSE)
write_pdb(bundle$posterior, pdbl, overwrite = FALSE)
write_pdb(bundle$reference_draws, pdbl, overwrite = FALSE)
```

Writing reference draws also writes both summary statistics. To write
them separately:

```{r, eval=FALSE}
write_pdb(bundle$reference_draws, pdbl, write_summary_statistics = FALSE)
write_pdb(bundle$summary_statistics$mean_value, pdbl)
write_pdb(bundle$summary_statistics$mean_squared_value, pdbl)
```

`mean_squared_value` is the mean of the squared draws, not a standard
deviation.

To check a posterior object, saved or not, use
`check_pdb_posterior(bundle$posterior, run_stan_code_checks = FALSE)`.
Use `check_pdb(pdbl)` to check a whole database.

## Import a fit for an existing posterior

When the posterior is already in the database and you only need to add
reference draws, use `import_reference_posterior_draws()`. It accepts
RStan and CmdStanR fits:

```{r, eval=FALSE}
pdbl <- pdb_local("/path/to/posteriordb/posterior_database")
imported_draws <- import_reference_posterior_draws(
  fit,
  posterior = "existing_data-existing_model",
  pdb = pdbl,
  write = TRUE,
  overwrite = FALSE
)
```

The importer always runs the full checks, and with `write = TRUE` it
writes only if they pass. It writes the draws and both summary
statistics (`write_summary_statistics = FALSE` skips the summaries) and
links them to the posterior. The existing data, model and posterior
records are not otherwise changed. Unlike the bundle writer, the import
is all or nothing: if any step fails, the original files are restored,
and you are warned if any could not be.

- The fit's unconstrained parameter counts must match the posterior's
  `dimensions`. The importer does not confirm that the fit used the
  model source and data linked to that posterior.
- A posterior with no linked reference draws gets the new link. A
  posterior already linked to different reference draws is refused.
- `include` and `exclude` work as described in [Choose
  variables](#choose-variables).
- With `write = FALSE`, or with `as_reference_posterior_draws()`, you
  get the checked draws back and nothing is written.

For CmdStanR fits, sample with `sig_figs = 18` and keep the model and
data files next to the fit. The parameter counts are computed by
CmdStan from the saved draws, and values rounded to the default six
significant figures can violate parameter constraints and be rejected.

## Add BibTeX references

`references` fields hold citation keys. A key must be in the database's
`bibliography/references.bib` before you check or write an object that
cites it. `append_reference()` adds entries in any of three forms:

```{r add-bibtex-references, eval=FALSE}
pdbl <- pdb_local("/path/to/posteriordb/posterior_database")

# BibTeX text: exactly one entry with a citation key.
append_reference(
  "@misc{my-study, title={My study}, year={2026}}", pdb = pdbl
)

# A .bib file: one or more entries.
append_reference("new-references.bib", pdb = pdbl)

# An R bibentry object: one or more entries.
ref <- utils::bibentry(
  "Misc", key = "my-software", title = "My model implementation",
  year = "2026"
)
append_reference(ref, pdb = pdbl)
```

Duplicate keys (ignoring case) and duplicate entries are rejected
without changing the bibliography. List the available keys with
`bibliography_keys(pdbl)`, and look one up with
`bibliography_entry("my-study", pdb = pdbl)`.

## Limitations

- `create_pdb_bundle()` accepts completed RStan HMC (NUTS) fits. It does
  not accept CmdStanR fits, optimization or variational results, or
  fits merged from several runs.
- The Stan source saved in the fit must be self-contained; source that
  uses `#include` is rejected.
- A variable must be saved completely to be kept. Zero-sized outputs
  cannot be kept; exclude them.
- The package cannot confirm that the data you supply produced the fit.
- Version information in a bundle describes the R session that built
  it. The Stan version requirement recorded for the model
  (`">=2.26.0"` by default) is not a record of the version used for
  sampling.
- If a `stanfit` was saved and reloaded in a new R session, the model
  may be recompiled to work out the parameter counts. The fit is not
  resampled. RStan may print
  `the number of chains is less than 1; sampling not done` while this
  happens; that is expected.

Use a separate copy of the database when experimenting with writes.

## Related tools

These functions work with reference draws but are not part of building a
bundle. Each has its own help page.

| Task | Function |
|------------------------------|------------------------------------------|
| Sample reference draws for several existing posteriors in one run | `sequential_batch_workflow()` |
| Recompute transformed parameters and generated quantities from saved parameter draws | `reconstruct_stan_output()`, `reconstruct_posterior_output()` |
| Work out a model's unconstrained parameter counts | `infer_posterior_dimensions()`, `infer_unconstrained_parameter_counts_from_fit()` |
| Link reference draws that are already saved to a posterior | `link_reference_posterior()` |
| Rename saved data, a model or a posterior | `rename_pdb()` |
| List saved reference draws or summaries | `reference_posterior_names(pdbl, type = "draws")` |

For example, to recompute `theta` for the eight-schools bundle from its
saved parameters:

```{r, eval=FALSE}
reconstructed <- reconstruct_stan_output(
  draws = bundle$reference_draws,
  model = bundle$model_code,
  data = bundle$data,
  variables = "theta"
)
reconstructed$draws
```

Deterministic outputs are reproduced to numerical precision when the
model and data match. Random generated quantities are drawn again, not
recovered.
