---
output:
  rmarkdown::md_document:
    variant: gfm
editor_options: 
  markdown: 
    wrap: 72
---

<!-- CREATE_PDB_BUNDLE.md is generated from CREATE_PDB_BUNDLE.Rmd. Please edit that file. -->

# Create and contribute a reference-draw bundle

`create_pdb_bundle()` builds linked data, model, posterior, and
reference-draw objects around an already sampled `rstan::stanfit`. The
result is an R list that you can inspect and check before writing any of
its objects. The function does not resample the fit or write files. It
supports the existing list-based workflow by default, and can reuse
existing data, model-code, or posterior objects (or retrieve them by
name from a database). It currently accepts RStan `stanfit` objects;
`import_reference_posterior_draws()` also accepts CmdStanR fits.

Choose the workflow based on whether the data, model, and posterior
records already exist:

- Use `create_pdb_bundle()` to construct linked objects from a fit. The
  standard workflow supplies the Stan data list plus `data_info` and
  `model_info`; the function builds everything in memory and leaves
  persistence to you. If records already exist, pass `data` as a
  `pdb_data` object or saved data name, and pass `model_code` and/or
  `posterior` as existing objects or saved names. A supplied posterior
  can provide its linked data and model code automatically. Existing
  model source must match the fit, and posterior links and unconstrained
  parameter counts must match.
- Use `import_reference_posterior_draws()` when you already have a
  PosteriorDB posterior and want to import a completed fit for it. This
  supports RStan and CmdStanR fits, checks them, and can write the draws
  and summaries while linking them to the existing posterior. It does
  not rewrite the existing data, model, or posterior payloads.

For example, to import into an existing posterior in a local database:

```{r, eval=FALSE}
pdbl <- pdb_local("/path/to/posterior_database")
imported_draws <- import_reference_posterior_draws(
  fit,
  posterior = "existing_data-existing_model",
  pdb = pdbl,
  write = TRUE,
  overwrite = FALSE
)
```

The importer always runs the full reference-draw checks. With
`write = TRUE`, it writes only if those checks pass. It writes both
`mean_value` and `mean_squared_value` summaries by default; set
`write_summary_statistics = FALSE` to skip them. A posterior already
linked to a different reference-draw object cannot be relinked by this
call. The importer selects every saved scalar column for each declared
base variable in posterior dimensions, and always retains all inferred
parameter-block variables. Declared counts for actual model parameters must
match the fit's unconstrained counts. Dimensions may also name saved derived
outputs, which have no independent unconstrained counts to compare. The
importer does not confirm that the fit used the exact model source and data
linked to that posterior.

To retain additional saved outputs, pass base names through `include`:

```{r, eval=FALSE}
imported_draws <- import_reference_posterior_draws(
  fit, posterior = "existing_data-existing_model", pdb = pdbl,
  include = c("theta", "log_lik"), write = FALSE
)
```

Every variable named in `posterior$dimensions`, and every parameter-block
variable inferred from the fit, is retained automatically even if omitted
from `include`. Additional outputs may be transformed
parameters or generated quantities; all their indexed columns are retained.
`exclude` removes additional selections and takes precedence over `include`,
but excluding a dimensions or parameter-block variable is an error. Unknown
names are errors. Without these arguments, only those required variables are
retained. This can retain more variables than older importer versions.
Diagnostics and written summaries cover the retained selection, not omitted
outputs. Existing dimensions are not changed. Required variables missing
from the saved fit cause an error. Retaining the complete parameter block
provides the inputs needed by the experimental reconstruction function, but
the importer does not itself reconstruct outputs or verify model/data identity.
Random generated quantities cannot generally be recovered exactly.

To build from saved data and model records while constructing the
missing posterior and reference draws, use:

```{r, eval=FALSE}
pdbl <- pdb_local("/path/to/posterior_database")
bundle <- create_pdb_bundle(
  fit,
  data = "existing_data",
  model_code = "existing_model",
  pdb = pdbl,
  check = FALSE
)
```

You can pass `pdb_data`, `pdb_model_code`, and `pdb_posterior` objects
instead of names. If only `posterior` is supplied, its linked data and
model code are obtained with `get_data()` and `model_code()`. The
ordinary list plus metadata workflow remains the default. When reusing
an object, omit its corresponding metadata list. If both are supplied,
the function warns and uses the existing object's metadata, ignoring the
corresponding list. When supplying an existing posterior, its
`dimensions` must use the fork's current meaning: one scalar
unconstrained-parameter count per model parameter. Older PosteriorDB
entries that store output shapes may fail validation and cannot be
passed through unchanged. The selected variables and their inferred
counts must also agree with the supplied posterior.

Prepare the fit and the exact named Stan input list used for sampling.
The eight schools example below follows the model in `CONTRIBUTING.md`:

```{r, eval=FALSE}
library(posteriordb)
library(rstan)

stan_data_file <- system.file("test_files/eight_schools.R", package = "posteriordb")
source(stan_data_file)

stan_file <- system.file(
  "test_files/eight_schools_noncentered.stan",
  package = "posteriordb"
)
stan_code <- paste(readLines(stan_file), collapse = "\n")
stan_model <- rstan::stan_model(model_code = stan_code)
fit <- rstan::sampling(
  stan_model,
  data = eight_schools,
  chains = 10,
  iter = 20000,
  warmup = 10000,
  thin = 10,
  seed = 4711,
  control = list(adapt_delta = 0.92)
)
```

Pass the same input list used for sampling, along with the required
names and titles for the data and model. `added_by` and `added_date` are
shared defaults for the objects in the bundle; a value in an individual
metadata list takes precedence. `posterior_info` also accepts the
optional character fields `urls`, `references`, and `keywords`; supplied
values are written to the posterior JSON. If posterior keywords are
omitted, the posterior JSON writes `"keywords": null`.

For model metadata, `framework = "stan"` is a convenience form: the
bundle constructor creates `model_implementations$stan` for you, using
`models/stan/<model name>.stan` and `stan_version = ">=2.26.0"`. You do
not need to build that nested list for the usual Stan workflow. If you
prefer to supply `model_implementations` yourself, that remains
supported; the Stan `model_code` path must match the path inferred from
the model name, and an explicit `stan_version` is retained. For example,
this shorthand is sufficient:

```{r, eval=FALSE}
model_info = list(
  name = "normal",
  title = "Normal model",
  framework = "stan"
)
```

`model_info$prior` is optional descriptive metadata, often a list of
keywords that points to prior information elsewhere in PosteriorDB. For
example, pass `prior = list(keywords = "stan_recommended_35dbfe6")` when
that reference is known. The bundle does not infer prior distributions
from the Stan source; if no prior metadata is supplied, the written
model info omits `prior`. The model writer does not emit a
`likelihood_code` entry. Optional model metadata is written only when
supplied; the writer does not add a `pymc` entry or a `pymc_version` by
default. If model keywords are omitted, the model info JSON writes
`"keywords": null`. The bundle API defaults the Stan implementation's
`stan_version` to `">=2.26.0"`. You can override this by supplying the
`stan_version` inside `model_info$model_implementations$stan`, while
keeping `model_code` equal to the inferred path, for example:

``` r
model_info = list(
  name = "normal",
  title = "Normal model",
  model_implementations = list(stan = list(
    model_code = "models/stan/normal.stan",
    stan_version = ">=2.35.0"
  ))
)
```

```{r, eval=FALSE}
bundle <- create_pdb_bundle(
  fit,
  data = eight_schools,
  added_by = "Stanislaw Ulam",
  data_info = list(
    name = "test_eight_schools_data",
    title = "A Test Data for the Eight Schools Model",
    description = "Treatment estimates and standard errors for eight schools."
  ),
  model_info = list(
    name = "test_eight_schools_model",
    title = "Test Non-Centered Model for Eight Schools",
    description = "A hierarchical model with a non-centered parameterization.",
    references = "rubin1981estimation"
  )
)
```

The constructor returns all four objects in memory:

```{r, eval=FALSE}
bundle$data
bundle$model_code
bundle$posterior
bundle$reference_draws
bundle$summary_statistics$mean_value
bundle$summary_statistics$mean_squared_value
```

The posterior embeds the data, model code, and reference draws, so its
getters work before persistence. `bundle$provenance` records how the
input data was obtained, the fit class, selected variables, and sampling
metadata. Passing a `pdb` connection attaches it to the objects; when a
saved object name is supplied, the connection is also used to retrieve
that object. Construction never writes database files.

If a serialized RStan fit has lost its compiled model pointer, bundle
construction recompiles the saved Stan source and uses the supplied data
to recover unconstrained parameter counts. This does not resample or
replace the fit's saved draws. RStan may print
`the number of chains is less than 1; sampling not done` during this
dimension-recovery step; that is expected.

## Choose when to compute diagnostics

`check = TRUE` is the default. It calculates the reference-draw
diagnostic report during bundle creation. The report is available as:

```{r, eval=FALSE}
bundle$diagnostics$metrics
bundle$diagnostics$status
bundle$diagnostics$failures
```

`status` contains one logical value per acceptance check. `failures`
identifies which checks failed and, where relevant, which variables or
chains were responsible. A failed check does not prevent bundle
creation: the bundle is returned with the failure recorded, so you can
inspect the report. Its reference draws cannot be written until they
pass.

When the reference-draw checks pass, the bundle also computes the
supported summary statistics (`mean_value` and `mean_squared_value`)
from those draws. They are returned in `bundle$summary_statistics`. An
unchecked bundle, or a bundle whose checks fail, has
`summary_statistics = NULL`; this prevents summaries from appearing
writeable before the required draw checks pass. The summary info objects
carry the same reference-posterior metadata as the draws, with the
summary-specific acceptance flags and summary-computation version added
by the existing summary-statistic constructor.

The acceptance checks require exactly 10,000 retained draws, at least
four chains, absolute lag-1 autocorrelation at most 0.05 for every
retained variable, R-hat at most 1.01 for every variable, E-FMI at least
0.2 for every chain, and no divergent transitions. Bulk and tail ESS are
also calculated and stored as diagnostic metrics, but ESS bounds do not
decide whether the draws pass.

Set `check = FALSE` to create the objects without calculating the
diagnostic metrics or evaluating acceptance checks. The bundle has
`diagnostics = NULL` and its reference draws have no acceptance flags.
Sampler diagnostics are retained in memory for a later check; sampling
is not repeated. Unchecked draws are not writable.

```{r, eval=FALSE}
unchecked <- create_pdb_bundle(
  fit,
  data = eight_schools,
  data_info = list(name = "test_eight_schools_data", title = "Eight schools data"),
  model_info = list(name = "test_eight_schools_model", title = "Eight schools model"),
  check = FALSE
)

is.null(unchecked$diagnostics)  # TRUE
info(unchecked$reference_draws)$checks_made  # NULL
```

To defer the check, pass the unchecked bundle to
`check_reference_posterior_draws()`. It returns an updated bundle with
the report and result attached:

```{r, eval=FALSE}
bundle <- check_reference_posterior_draws(unchecked)
bundle$diagnostics$status
bundle$diagnostics$failures
```

The bundle workflow currently supports checking the full set of
acceptance criteria during construction or checking that full set later.
It does not provide a way to select one criterion for a bundle check.
For investigating an individual check on a sampled RStan or CmdStanR fit
before importing it, use `reference_draw_diagnostics()` and
`passes_reference_draw_checks()`. These functions accept a `checks`
selector such as `"mean_lag1_ac"`; passing one check does not establish
full reference-draw acceptance or make a draw object writable.

```{r, eval=FALSE}
report <- reference_draw_diagnostics(fit, checks = "mean_lag1_ac")
report$metrics
report$thresholds
report$status
report$failures
passes_reference_draw_checks(fit, checks = "mean_lag1_ac")
```

## Choose variables and write the objects

By default, the bundle includes saved parameters, transformed
parameters, and generated quantities, excluding `lp__`: this is the
`include = NULL` default. Use `include = character(0)` to retain only the
complete parameter block. In R, `c()` is `NULL`, so `include = c()` selects
all outputs rather than parameter-block-only draws. Use `include` or
`exclude` with base variable names to select additional outputs. For example,
`include = c("mu", "tau")` keeps those variables **and every inferred
parameter-block variable**. `include = "theta"` can additionally retain
all elements of a transformed parameter or generated quantity. Excluding
an inferred parameter-block variable is an error; derived outputs can be
excluded. The same retained selection is used for immediate or deferred
diagnostics and summary statistics. The posterior's `dimensions` entries
always describe the complete inferred parameter block and are unaffected
by selection of derived outputs. These entries record unconstrained
parameter counts, which can differ from the number or shape of saved
output columns (for example, a constrained simplex has one fewer
unconstrained coordinate than output elements).

The `dimensions` map describes the model parameters that have
unconstrained coordinates; it is not a complete inventory of every
variable saved in the draws. Transformed parameters and generated
quantities can be present in the draw payload without appearing in this
map. Use `include` and `exclude` to choose which saved variables are
retained and used for the bundle's draw diagnostics. In particular,
exclude a generated quantity if it is not useful for diagnosing the fit.

For an existing fit, the package can infer the named unconstrained
counts directly from either supported fit class:

```{r, eval=FALSE}
rstan_dimensions <- infer_unconstrained_parameter_counts_from_fit(rstan_fit)
cmdstanr_dimensions <- infer_unconstrained_parameter_counts_from_fit(cmdstanr_fit)
```

The same helper dispatches on fit class and returns one count per model
parameter. If there is no fit object, `infer_posterior_dimensions()` can
infer counts from Stan source and data; see
`?infer_posterior_dimensions` for its backend-specific compile behavior.

You can write the whole bundle with one call. If diagnostics have not
been computed yet, this checks the draws first. New data, model, and
posterior objects are written either way; records reused from the
destination database are left in place. Reference draws and their
summary statistics are written only if all acceptance checks pass.
Before writing, the bundle writer checks every planned output path,
including info files, payloads, and summaries:

```{r, eval=FALSE}
pdbl <- pdb_local()
write_result <- write_pdb(bundle, pdbl, overwrite = FALSE)
write_result$written
write_result$reused
write_result$collisions
write_result$overwritten
write_result$reference_draws_written
write_result$skipped_reason
```

When a draw check fails, `write_result$bundle$diagnostics$failures`
describes the failed checks. A diagnostic error is recorded in
`write_result$diagnostic_error`; in either case, the three non-draw
components are written or reused, while draws and summaries are skipped.

If a planned file already exists, `overwrite = FALSE` stops before
writing any bundle files and reports all colliding paths.
`overwrite = TRUE` still runs the same preflight, reports those paths in
`write_result$collisions`, and allows replacement of newly constructed
components. Data, model, or posterior objects reused from the
destination database are instead left in place and reported in
`write_result$reused`; the bundle writer never overwrites those records.
If a reused object comes from another database and the target has no
files with that name, it can be copied to the target. Partial or
ambiguous same-name files for a reused object stop the operation
regardless of `overwrite`.

After preflight, the writes remain sequential in data, model, posterior,
then reference-draw and summary order. Preflight prevents file-name
collisions from causing a partial write, but a later serialization or
filesystem error can still leave earlier successful writes in place. To
add accepted draws to a database that already has the matching data,
model, and posterior, write only the reference-draw object:

```{r, eval=FALSE}
write_pdb(bundle$reference_draws, pdbl, overwrite = FALSE)
```

The associated posterior JSON must already exist. This individual draw
write also writes both supported summary statistics by default. Use the
whole-bundle writer when the destination does not already contain
conflicting objects, or when replacing them with `overwrite = TRUE` is
intended. That option replaces files at the bundle's destination names;
it does not merge metadata or protect against replacing a different
object's files that use those same names.

You can also write the individual objects manually when you want to
control the steps separately. The individual reference-draw writer
rejects unchecked or failed draws and requires the associated posterior
JSON to exist first:

```{r, eval=FALSE}
assert_checked_reference_posterior_draws(bundle$reference_draws)
write_pdb(bundle$data, pdbl, overwrite = FALSE)
write_pdb(bundle$model_code, pdbl, overwrite = FALSE)
write_pdb(bundle$posterior, pdbl, overwrite = FALSE)
write_pdb(bundle$reference_draws, pdbl, overwrite = FALSE)
```

By default, the reference-draw `write_pdb()` method also computes and
writes the supported summary statistics (`mean_value` and
`mean_squared_value`) using their existing constructors and writer
methods. Data, model, reference-draw, and summary-statistic payloads are
written with their info files; summary statistics are saved under
`reference_posteriors/summary_statistics/<type>/`. The posterior JSON
records the links and dimensions. `overwrite = FALSE` is the default and
causes an error if a destination file already exists. Each write happens
separately, so successful earlier writes remain if a later write fails.
Set `overwrite = TRUE` only when replacing existing files is intended.

If the associated posterior JSON is missing, writing reference draws
stops before writing any reference-draw or summary-statistic files. This
ensures the saved reference posterior is linked from an existing
posterior entry.

To save only the draws in that call, use
`write_summary_statistics = FALSE`. You can then write the summary
objects from the checked bundle individually:

```{r, eval=FALSE}
write_pdb(
  bundle$reference_draws,
  pdbl,
  overwrite = FALSE,
  write_summary_statistics = FALSE
)
write_pdb(bundle$summary_statistics$mean_value, pdbl, overwrite = FALSE)
write_pdb(bundle$summary_statistics$mean_squared_value, pdbl, overwrite = FALSE)
```

The summary payload field names follow the existing PosteriorDB format:
`mean_value` contains `names`, `mean_value`, and `mcse_mean`;
`mean_squared_value` contains `names`, `mean_squared_value`, and
`mcse_mean`. The latter is the mean of squared draws, not a standard
deviation. Their info JSON uses the reference-draw metadata shape, with
`versions$r_summary_statistic` identifying the package version used to
compute the summaries.

For further details about accepted metadata and arguments, see
`?create_pdb_bundle`. The constructor requires the actual named Stan
input list for the default workflow. Alternatively, pass an existing
`pdb_data` object or its saved name, or supply a posterior whose linked
data can be retrieved. In all cases, the function cannot establish that
the supplied data was the data used to produce the fit.
