---
output:
  rmarkdown::md_document:
    variant: gfm
---

<!-- CREATE_PDB_BUNDLE.md is generated from CREATE_PDB_BUNDLE.Rmd. Please edit that file. -->

Create and contribute a reference-draw bundle
============================================

`create_pdb_bundle()` builds linked data, model, posterior, and reference-draw
objects around an already sampled `rstan::stanfit`. The result is an R list
that you can inspect and check before writing any of its objects. The function
does not compile or sample a model, read from a database, or write files. The
current implementation supports RStan `stanfit` objects.

Prepare the fit and the exact named Stan input list used for sampling. The
eight schools example below follows the model in `CONTRIBUTING.md`:

```{r, eval=FALSE}
library(posteriordb)
library(rstan)

stan_data_file <- system.file("test_files/eight_schools.R", package = "posteriordb")
source(stan_data_file)

stan_file <- system.file(
  "test_files/eight_schools_noncentered.stan",
  package = "posteriordb"
)
stan_code <- paste(readLines(stan_file), collapse = "\n")
stan_model <- rstan::stan_model(model_code = stan_code)
fit <- rstan::sampling(
  stan_model,
  data = eight_schools,
  chains = 10,
  iter = 20000,
  warmup = 10000,
  thin = 10,
  seed = 4711,
  control = list(adapt_delta = 0.92)
)
```


Pass the same input list used for sampling, along with the required names and
titles for the data and model. `added_by` and `added_date` are shared defaults
for the objects in the bundle; a value in an individual metadata list takes
precedence.

```{r, eval=FALSE}
bundle <- create_pdb_bundle(
  fit,
  data = eight_schools,
  added_by = "Stanislaw Ulam",
  data_info = list(
    name = "test_eight_schools_data",
    title = "A Test Data for the Eight Schools Model",
    description = "Treatment estimates and standard errors for eight schools."
  ),
  model_info = list(
    name = "test_eight_schools_model",
    title = "Test Non-Centered Model for Eight Schools",
    description = "A hierarchical model with a non-centered parameterization.",
    references = "rubin1981estimation"
  )
)
```


The constructor returns all four objects in memory:

```{r, eval=FALSE}
bundle$data
bundle$model_code
bundle$posterior
bundle$reference_draws
bundle$summary_statistics$mean_value
bundle$summary_statistics$sd
```


The posterior embeds the data, model code, and reference draws, so its getters
work before persistence. `bundle$provenance` records how the input data was
obtained, the fit class, selected variables, and sampling metadata. Passing a
`pdb` connection only attaches it to the objects; it does not read or write
database files.

Choose when to compute diagnostics
----------------------------------

`check = TRUE` is the default. It calculates the reference-draw diagnostic
report during bundle creation. The report is available as:

```{r, eval=FALSE}
bundle$diagnostics$metrics
bundle$diagnostics$status
bundle$diagnostics$failures
```


`status` contains one logical value per acceptance check. `failures` identifies
which checks failed and, where relevant, which variables or chains were
responsible. A failed check does not prevent bundle creation: the bundle is
returned with the failure recorded, so you can inspect the report. Its
reference draws cannot be written until they pass.

When the reference-draw checks pass, the bundle also computes the supported
summary statistics (`mean_value` and `sd`) from those draws. They are returned
in `bundle$summary_statistics`. An unchecked bundle, or a bundle whose checks
fail, has `summary_statistics = NULL`; this prevents summaries from appearing
writeable before the required draw checks pass. The summary info objects carry
the same reference-posterior metadata as the draws, with the summary-specific
acceptance flags and summary-computation version added by the existing
summary-statistic constructor.

The acceptance checks require exactly 10,000 retained draws, at least four
chains, absolute lag-1 autocorrelation at most 0.05 for every retained
variable, R-hat at most 1.01 for every variable, E-FMI at least 0.2 for every
chain, and no divergent transitions. Bulk and tail ESS are also calculated and
stored as diagnostic metrics, but ESS bounds do not decide whether the draws
pass.

Set `check = FALSE` to create the objects without calculating the diagnostic
metrics or evaluating acceptance checks. The bundle has `diagnostics = NULL`
and its reference draws have no acceptance flags. Sampler diagnostics are
retained in memory for a later check; sampling is not repeated. Unchecked draws
are not writable.

```{r, eval=FALSE}
unchecked <- create_pdb_bundle(
  fit,
  data = eight_schools,
  data_info = list(name = "test_eight_schools_data", title = "Eight schools data"),
  model_info = list(name = "test_eight_schools_model", title = "Eight schools model"),
  check = FALSE
)

is.null(unchecked$diagnostics)  # TRUE
info(unchecked$reference_draws)$checks_made  # NULL
```


To defer the check, pass the unchecked bundle to
`check_reference_posterior_draws()`. It returns an updated bundle with the
report and result attached:

```{r, eval=FALSE}
bundle <- check_reference_posterior_draws(unchecked)
bundle$diagnostics$status
bundle$diagnostics$failures
```


The bundle workflow currently supports checking the full set of acceptance
criteria during construction or checking that full set later. It does not
provide a way to select one criterion for a bundle check.

Choose variables and write the objects
--------------------------------------

By default, the bundle includes saved parameters, transformed parameters, and
generated quantities, excluding `lp__`. Use `include` or `exclude` with base
variable names to select variables. For example, `include = c("mu", "tau")`
keeps all saved scalar elements of those variables.

Only write reference draws after all acceptance checks pass. The assertion
below stops before writing if they are unchecked or failed. Point `pdb_local()`
to the local database you intend to update, then write the four objects
individually:

```{r, eval=FALSE}
assert_checked_reference_posterior_draws(bundle$reference_draws)

pdbl <- pdb_local()
write_pdb(bundle$data, pdbl, overwrite = FALSE)
write_pdb(bundle$model_code, pdbl, overwrite = FALSE)
write_pdb(bundle$reference_draws, pdbl, overwrite = FALSE)
write_pdb(bundle$summary_statistics$mean_value, pdbl, overwrite = FALSE)
write_pdb(bundle$summary_statistics$sd, pdbl, overwrite = FALSE)
write_pdb(bundle$posterior, pdbl, overwrite = FALSE)
```


Each call uses the existing `write_pdb()` method for that object. Data, model,
reference-draw, and summary-statistic payloads are written with their info
files; summary statistics are saved under
`reference_posteriors/summary_statistics/<type>/`. The posterior JSON records
the links and dimensions. `overwrite = FALSE` is the default and causes an
error if a destination file already exists. Each write happens separately, so
successful earlier writes remain if a later write fails. Set `overwrite = TRUE`
only when replacing existing files is intended.

For further details about accepted metadata and arguments, see
`?create_pdb_bundle`. The constructor requires the actual named Stan input
list; it cannot establish that supplied data was the data used to produce the
fit.
